---
title: "About"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{About}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
description: >
  A package for germplasm curators to identify regions to sample material from. 
---

## `safeHavens`

This package enables germplasm curators to identify and communicate areas of interest to collection teams, supporting the targeting of new accessions.
For common species, it offers several options for curators to choose where to collect each plant type.
For rare species, the package offers an additional sampling approach.
The package can be integrated into existing workflows. It reads accession data from databases or Excel, compares them to occurrence data, and outputs results in tabular formats (CSV, database, Excel) and spatial formats (geopackages, shapefiles). These results can be shared with collection teams and used on handheld devices such as Qfield (free, preferred) or ESRI products (paid, legacy).
Curators gain spatial capabilities through the package without requiring GIS expertise or paid software.

### 'common' species sampling approaches

Each approach is grounded in genetics and ecological theory, and aligns with landscape genetics and genomics concepts.
All functions are designed to capture geographic and environmental variation across a species' range.
To this end, they evolve from simply partitioning a species' range into `n` equal or regularly spaced areas to modelling isolation by distance, resistance, and environment, or sampling from existing spatial data such as ecoregions or seed transfer zones.
Distance functions are based on Sewall Wright's (1943) idea that plants growing closer together are more alike than those farther apart.
All approaches in the package are a priori (based on theoretical reasoning rather than empirical observation), except for the EnvironmentalBasedSample function. This function uses a species distribution model (SDM, a tool that predicts where a species is likely to occur based on environmental variables) to inform sampling locations and is based on the concept of Isolation by Environment (Wang & Bradburd, 2014). However, this approach lacks field validation.
None of the sampling approaches is meant to replace genetic diversity-based sampling. However, such data is rare, and gathering it often means missing opportunities for opportunistic seed collections.
The methods have trade-offs in computational and environmental complexity. All are designed to run on standard computers. With the faster methods, a user can process a couple of thousand species overnight.
The table below presents the currently implemented sampling scheme and the user-facing function associated with it.

|        Function           |              Description               | Comp.| Envi.|
|---------------------------|----------------------------------------|------|------|
| `PointBasedSample`        | Creates points to make grids over area |  L   |   L  |
| `EqualAreaSample`         | Breaks area into similar size pieces   |  L   |   L  |
| `OpportunisticSample`     | Using PBS with existing records        |  L   |   L  |
| `IBDBasedSample`          | Breaks species range into clusters     |  H   |   L  |
| `IBRSurface`              | Breaks species range into clusters     |  H   |   M  |
| `PolygonBasedSample`      | Using existing ecoregions or STZs      |  L   |   H  |
| `EnvironmentalBasedSample`| Uses correlations from SDM to sample   |  H   |   H  |
| `KMedoidsBasedSample`     | For rare species with known occurrences|  M   |   M  |
| `PosteriorCluster`        | Uses posteriors from an SDM to cluster |  H   |   H  |

### geographic distance 
The functions `PointBasedSample` and `EqualAreaSample` partition the species' range into geographic clusters of similar size.
The `OpportunisticSample` method is a variant of `PointBasedSample` that uses existing collection records to guide the sampling scheme, designing additional samples around current records.
`OpportunisticSample` uses the same logic as `PointBasedSample`. First, it removes areas within a specified distance of existing collection records—known as buffer zones—then fills the remaining gaps with new sampling locations.
Use `PointBasedSample` and `EqualAreaSample` to establish a new collection strategy, or `OpportunisticSample` to augment an existing strategy.

`IBDBasedSample` also relies on geographic distance, but instead of using the continuous nature of space, it focuses on discontinuities. It uses distance matrices (tables showing the distance between all pairs of points) and clustering (grouping similar points) to identify portions of the species' range that are closer to each other than to other parts of the range.
This method incurs a slightly higher computational cost than previous geographic distance approaches.

### using regions 

A polygon-based sample may be most common in North America. 
In various forms, it is used in major germplasm banking projects in the Midwest and Southeastern United States.
It also shapes how native seed collection coordinators are structured in the West.
This method uses environmental variation to guide seed collections. 
Different ecoregions act as stratification agents.
Generally, these regions exhibit continuous environmental transitions that the species face. 
Populations in different regions adapt uniquely.
It can be used with either ecoregion- or seed-transfer-zone-based data.
The method relies on existing spatial data products. These may or may not match the species' ecology.

### resistance distance 
The `IBRSurface` workstream lets users create a general cost surface.
Least-cost paths (the easiest routes between two locations, based on the cost surface) are calculated and grouped for occupied sections of the species' range.
The least-cost paths between occupied parts of the range are calculated and then clustered.
Spatial data from the end of this process can be passed into `PolygonBasedSample` if more than one collection per zone is desired.

### environmental distance 

`EnvironmentalBasedSample` is both the most computationally expensive and the most environmentally explicit.
This function fits a Species Distribution Model (SDM) using a generalised linear model (glmnet, a regularised regression method). It clusters populations based on environmental variables, observed species locations, and the spatial arrangement of those populations.
On paper, this draws together all aspects of the above functions. However, this approach has not been empirically tested.

Spatial data from the end of this process can be passed into `PolygonBasedSample` if more than one collection per zone is desired.

## rare species sampling approach

All of the above methods are designed for common species, that is, species with more than 50 or 100 unique occurrence records.Each method returns a polygon geometry that gives guidelines for the general regions where sampling should occur.
However, for rare species, where the occurrence of individual populations are tracked, and which are generally collected along maternal lines, requiring considerably more field effort to gather seed, a supplemental approach exists that suggests the priority in which individual populations (‘points’) can be sampled. 

`KMedoidsBasedSample` uses geographic distances alone or, optionally, a distance matrix derived from key environmental variables. It suggests which populations to prioritise for seed collections.
This method is based on the concept of maximising the dispersion of selected points across geographic and environmental spaces to best capture the overall variation in the species' range.

## BAYESIAN GOES HERE

### installation 

`safeHavens` can be installed directly from GitHub, using either devtools or remotes.
The software relies on other packages that may require additional system dependencies, such as rgeos and rgdal.
It also requires GDAL, PROJ, and GEOS to be installed on your system.
This can occasionally be finicky to install, but once you have it, you effectively have access to all open-source geospatial software.
Users of Bayesian approaches must have Stan and cmdstanr installed. These also require C++ toolkits.
Installation may require additional configuration time.
`safeHavens` can be installed from GitHub with `remotes`.

```{r Install Package, message=FALSE, eval = F}
remotes::install_github('sagesteppe/safeHavens') 
```

An overview of the functionality in the package is below. 

```{r echo = F, fig.width=6, fig.height=3}
DiagrammeR::mermaid("
%%{init: {'theme':'dark', 'themeVariables': { 
    'primaryTextColor':'#ffffff', 
    'lineColor':'#ffffff',
    'lineWidth':'24px',
    'fontSize':'30px', 
    'fontFamily':'Arial'}}}%%
flowchart LR
A[/Species Occurrence Data/]
B[/Environmental Covariates/]

A --> D(PointBasedSample)
A --> E(EqualAreaSample)
A --> F(OpportunisticSample)
A --> G(IBDBasedSample)
A --> J[elasticSDM]
A --> K[bayesianSDM]
A --> L(KMedoidsBasedSample)
B -.-> L
B --> J
B --> K
D --> N[PrioritizeSample]
E --> N
F --> N
G --> N
A --> U[populationResistance]
B --> H[buildResistanceSurface]
H --> U
U --> V[IBRSurface]
V --> M[PolygonBasedSample]
M --> N
J --> O{postProcessSDM}
K --> O
O --> P{RescaleRaster}
O --> S{RescaleRasterBayes}
P --> Q(EnvironmentalBasedSample)
Q --> R[PredictiveProvenance]
Q --> M
R --> M
S --> T[PosteriorCluster]
T --> R
classDef geoColor fill:#d95f02,color:#FFFFFF
classDef polyColor fill:#66a61e,color:#FFFFFF
classDef envColor fill:#1b9e77,color:#FFFFFF
classDef dataColor fill:#a6761d,color:#FFFFFF
classDef decisionColor fill:#7570b3,color:#FFFFFF
classDef rareColor fill:#e6ab02,color:#FFFFFF
classDef ibrColor fill:#e7298a,color:#FFFFFF
class D,E,F,G geoColor
class H,M polyColor
class J,K,O,P,S,T,U,Q,R envColor
class A,B dataColor
class N decisionColor
class L rareColor
class H,U,V ibrColor
")
```