---
title: "Bayesian Approaches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Approaches}
  `%\VignetteEngine{knitr::rmarkdown_notangle}`
  %\VignetteEncoding{UTF-8}
description: >
  Using Bayesian approaches for EnvironmentalBasedSample.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

# Species Distribution Modelling

Please refer to all frequentist methods before viewing this vignette. 
The rationale is roughly similar, but this approach provides better estimates of uncertainty for the processing. 

## Background

This vignette details the steps required to create a bayesian Species Distribution Model (SDM) using the functions provided in `safeHavens`, which form a wrapper around brms::brms(). 
The SDM is then post-processed to create a binary raster map of suitable and unsuitable habitat, that is used to rescale the environmental predictor variables according to the parameter posteriors for the model. 
These parameter posteriors are then used in a clustering algorithm to partition the species range into environmentally distinct regions for germplasm sampling.

The goal of most SDM's is to create a model that accurately predicts where a species will be located in environmental space and can then be predicted in geographic space.
The goal of these models is to understand the degree and direction to which various environmental features correlate with a species observed range. 

## About 

While the authors are prone to creating many 'advanced' SDMs models and in-house pipelines for developing them, we think that the juice is not worth the squeeze and rely on a few powerhouse R packages to do the heavy lifting.
The main packages used are `dismo`, `caret`, `brms`, `CAST`, and `terra`. 

# Steps to create an SDM

### prep data 

The below steps are all present at the start of the 'Predictive Provenance' vignette. 
We will 'hide' some of the processing steps to keep this vignette entry relatively short. 

```{r load libraries, message =F, warning = F}
#remotes::install_github('sagesteppe/safeHavens@bayes')
library(safeHavens)
library(terra)
library(geodata)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```


```{r Download species occurrence data, echo = FALSE}
cols = c('decimalLatitude', 'decimalLongitude', 'dateIdentified', 'species', 'acceptedScientificName', 'datasetName', 
  'coordinateUncertaintyInMeters', 'basisOfRecord', 'institutionCode', 'catalogNumber')

## download species data using scientificName, can use keys and lookup tables for automating many taxa. 
hemi <- rgbif::occ_search(scientificName = "Helianthella microcephala")
hemi <- hemi[['data']][,cols]  |>
  drop_na(decimalLatitude, decimalLongitude) |> # any missing coords need dropped. 
  distinct(decimalLatitude, decimalLongitude, .keep_all = TRUE) |> # no dupes can be present
  st_as_sf(coords = c( 'decimalLongitude', 'decimalLatitude'), crs = 4326, remove = F) 
```


```{r prep basemap, echo = F}
western_states <- spData::us_states |> ## for making a quick basemap. 
  dplyr::filter(NAME %in% 
    c('Utah', 'Arizona', 'Colorado', 'New Mexico', 'Wyoming', 'Nevada', 'Idaho', 'California')) |>
  dplyr::select(NAME, geometry) |>
  st_transform(4326)

bb <- st_bbox(
  c(
    xmin = -114.5, 
    xmax = -106, 
    ymax = 42, 
    ymin = 33.5
    ),
    crs = st_crs(4326)
    )

western_states <- st_crop(western_states, bb)

bmap <- ggplot() + 
    geom_sf(data = western_states) + 
    geom_sf(data = hemi) +
    theme_minimal() +
    coord_sf(
        xlim = c(bb[['xmin']], bb[['xmax']]), 
        ylim = c(bb[['ymin']], bb[['ymax']])
        )  +
    theme(
      axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank()
      )

bmap +
    labs(title = 'Helianthella microcephala\noccurrence records')
```

```{r remove cols and western states, echo = F}
rm(cols, western_states)
```

```{r download data, echo = FALSE}
# Download WorldClim bioclim at ~10 km
bio_current <- worldclim_global(var="bioc", res=2.5)

# Crop to domain 
bbox <- ext(bb)

bio_current <- crop(bio_current, bbox)
#bio_future <- crop(bio_future, bbox)
```

```{r standardize raster layer names, echo = FALSE}
simplify_names <- function(x){
    paste0('bio_', sprintf("%02d", as.numeric(gsub('\\D+','', names(x)))))
}

names(bio_current) <- gsub('^.*5m_', '', names(bio_current))
#names(bio_future) <- gsub('^.*2060_', '', names(bio_future))

names(bio_current) <- simplify_names(bio_current)
#names(bio_future) <- simplify_names(bio_future)

rm(simplify_names)
```


## fit the model
Here we fit as SDM using `bayesianSDM`. 
For arguments it requires occurrence data `x`, a raster stack of `predictors`, a `quantile_v` that is an offset used to create pseudo-absence data, and a `planar_proj` used to calculate distances between occurrence data and possible pseudo-absence points. 

```{r Create Species Distribution Model, message=F, warning=F}
hemi <- select(hemi) 

sdModel <- bayesianSDM( 
  x = hemi,
  pca_predictors = FALSE,
  predictors = bio_current,
  quantile_v = 0.025,
  planar_proj = 5070, 
  resample = TRUE,
  feature_selection = 'ffs',
  min_ffs_var = 5, 
  fact = 3, 
  silent = 2 # print fewer messages during model fit.
  )
```

Under the hood this function uses `brms`, which dispatches to Stan, to fit a bayesian hierarchical model (GLMM), where the distances between occurrence records are the fixed effect, useful to reduce the effects of spatial autocorrelation to increase interpret-ability of the models parameters.  

### explore the output

Plot the predictions 
```{r Explore SDM output - Different alpha}
sdm_td <- sdModel$TrainData
plot(sdModel$RasterPredictions)
points(vect(sdm_td[sdm_td$occurrence==1,]),col = 'black')
points(vect(sdm_td[sdm_td$occurrence==0,]),col = 'red')

plot(sdModel$RasterPredictions_sd)
points(vect(sdm_td[sdm_td$occurrence==1,]),col = 'black')
points(vect(sdm_td[sdm_td$occurrence==0,]),col = 'red')
```

The brms:: model summary can be accessed from the sdModel list. 
```{r}
sdModel$Model
```

THINGS TO LOOK OUT FOR: 

Diagnostics can be accessed from a list, and the stan website has some information on them... 

```{r check diagnostics}
sdModel$Diagnostics
```


The confusion matrix can be accessed from the test and train splits. 

```{r Explore SDM output - Confusion Matrix}
sdModel$ConfusionMatrix
```

Here we can see how our selected model works on predicting the state of test data.
Note these accuracy results are slightly higher than those from the CV folds. 
This is not a bug, CV folds are testing on their own holdouts, while this is a brand new set of holdouts. 
The main reason the confusion matrix results are likely to be higher is due to spatial auto-correlation which are not address in a typical 'random split' of test and train data.
In order to minimize the effects of spatial-autocorrelation on our model we use `CAST` under the hood which allows for spatially informed cross validation. 

Consider the output from CVStructure to be a bit more realistic. 

### binarize the output

The bayesian sdm can be post-processed using the same function as the `elasticSDM`.

```{r Threshold the SDM output}
threshold_rasts <- PostProcessSDM(
  rast_cont = sdModel$RasterPredictions, 
  test = sdModel$TestData,
  train = sdModel$TrainData,
  planar_proj = 5070,
  thresh_metric = 'equal_sens_spec', 
  quant_amt = 0.2
  )
knitr::kable(threshold_rasts$Threshold)
```

We can compare the results of applying this function side by side using the output from the function. 

```{r Compare Threshold Results}
terra::plot(threshold_rasts$FinalRasters)
```

### rescale predictor variables
 
The `bayesianSDM` requires a different function from `elasticSDM` for creating weighted surfaces, with the suffix `RescaleRasters_bayes`. 

```{r Rescale Predictor Variables}
rr <- RescaleRasters_bayes(
  model = sdModel$Model,
  predictors = sdModel$Predictors, 
  training_data = sdModel$TrainData, 
  pred_mat = sdModel$PredictMatrix,
  include_uncertainty = TRUE
  )

terra::plot(rr$RescaledPredictors)
```

We can also show the variables that contributed the most to the sdm models predictions. 

```{r Show beta Coefficients from model}
knitr::kable(rr$BetaCoefficients[ order(rr$BetaCoefficients$Estimate), ])
```


```{r posterior clustering}
obby <- PosteriorCluster(
  model = sdModel$Model,
  predictors = subset(rr$RescaledPredictors, 'coef_uncertainty', negate = TRUE),
  f_rasts = threshold_rasts$FinalRasters,
  lyr = 'Supplemented',
  pred_mat = sdModel$PredictMatrix,
  training_data = sdModel$Train,
  planar_proj = 5070
)
```

```{r plot stability raster with sample points}
plot(obby$StabilityRaster)
points(obby$SamplePoints)
```

```{r show posterior clusters}
ggplot(obby$Geometry) + 
  geom_sf(aes(fill = factor(ID)))
```

```{r view top 3 ranks side by side}
plot(obby$SummaryRaster$consensus)
plot(obby$RankRaster$rank2_final)
plot(obby$RankRaster$rank3_final)
```

### save results 

```{r Save SDM results, eval = F}
bp <- '~/Documents/assoRted/StrategizingGermplasmCollections'

writeSDMresults(
  cv_model = sdModel$CVStructure, 
  model = sdModel$Model, 
  cm = sdModel$ConfusionMatrix, 
  coef_tab = rr$BetaCoefficients, 
  f_rasts = threshold_rasts$FinalRasters,
  thresh = threshold_rasts$Threshold,
  file.path(bp, 'results', 'SDM'), 'Bradypus_test')

# we can see that the files were placed here using this. 
list.files( file.path(bp, 'results', 'SDM'), recursive = TRUE )
```

### wrapping up 

While the bayesian hierarchical GLM does not produce excellent predictions of suitable habitat, it has several advantages. 
  1) It can incorporate spatial smoothers (splines) from mcgv to reduce the effects of spatial-autocorrelation on model parameter estimates. 
  2) It provides uncertainty around the contribution of each selected predictor (i.e. many predictors can be weighed differently to get similar results)
  3) 

```{r Clean up SDM variables, warning=FALSE, echo = FALSE}
rm(rr, predictors, files, sdModel, bp)
```
