---
title: "Bayesian Approaches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Approaches}
  `%\VignetteEngine{knitr::rmarkdown}`
  %\VignetteEncoding{UTF-8}
description: >
  Using Bayesian approaches for EnvironmentalBasedSample.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

# Species Distribution Modelling

Before proceeding, please refer to all other vignettes, as they provide valuable foundational context. 
While the rationale here is roughly the same, this approach focuses on improved uncertainty estimation during processing.

## Background

This vignette details the steps required to create a Bayesian Species Distribution Model (SDM) using functions from `safeHavens`, which serve as wrappers for brms::brms(). Building on the foundational rationale mentioned previously, the following information guides you through the Bayesian-specific setup.
The SDM is then post-processed to create a binary raster map of suitable and unsuitable habitat, which is used to rescale the environmental predictor variables based on the model's parameter posteriors.
These parameter posteriors are then used in a clustering algorithm to partition the species range into environmentally distinct regions for germplasm sampling.

The goal of most SDMs is to create a model that accurately predicts where a species will be located in environmental space and can then be predicted in geographic space.
The goal of these models is to understand the degree and direction to which various environmental features correlate with a species' observed range.

## About

In this section, we use Bayesian hierarchical models to estimate coefficient uncertainty. This enables clustering climatically similar areas using an SDM approach, linking the previously described methodology to specific analytical outcomes.

### prep data 

The data preparation steps described below are also found at the start of the 'Predictive Provenance' vignette. Recall these foundational steps as we transition into Bayesian-specific applications.
To keep this vignette concise, some intermediate processing steps will be 'hidden.' This ensures the focus remains on the distinct Bayesian approach discussed in the sections above.

```{r load libraries, message =F, warning = F}
library(safeHavens)
library(terra)
library(geodata)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```

```{r Download species occurrence data, echo = FALSE}
cols = c('decimalLatitude', 'decimalLongitude', 'dateIdentified', 'species', 'acceptedScientificName', 'datasetName', 
  'coordinateUncertaintyInMeters', 'basisOfRecord', 'institutionCode', 'catalogNumber')

## download species data using scientificName, can use keys and lookup tables for automating many taxa. 
hemi <- rgbif::occ_search(scientificName = "Helianthella microcephala")
hemi <- hemi[['data']][,cols]  |>
  drop_na(decimalLatitude, decimalLongitude) |> # any missing coords need dropped. 
  distinct(decimalLatitude, decimalLongitude, .keep_all = TRUE) |> # no dupes can be present
  st_as_sf(coords = c( 'decimalLongitude', 'decimalLatitude'), crs = 4326, remove = F) 
```

```{r prep basemap, echo = F, warning=F}
western_states <- spData::us_states |> ## for making a quick basemap. 
  dplyr::filter(NAME %in% 
    c('Utah', 'Arizona', 'Colorado', 'New Mexico', 'Wyoming', 'Nevada', 'Idaho', 'California')) |>
  dplyr::select(NAME, geometry) |>
  st_transform(4326)

bb <- st_bbox(
  c(
    xmin = -114.5, 
    xmax = -106, 
    ymax = 42, 
    ymin = 33.5
    ),
    crs = st_crs(4326)
    )

western_states <- st_crop(western_states, bb)

bmap <- ggplot() + 
    geom_sf(data = western_states) + 
    geom_sf(data = hemi) +
    theme_minimal() +
    coord_sf(
        xlim = c(bb[['xmin']], bb[['xmax']]), 
        ylim = c(bb[['ymin']], bb[['ymax']])
        )  +
    theme(
      axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank()
      )

bmap +
    labs(title = 'Helianthella microcephala\noccurrence records')
```

```{r remove cols and western states, echo = F}
rm(cols, western_states)
```

```{r download data, echo = FALSE}
# Download WorldClim bioclim at ~10 km
bio_current <- worldclim_global(var="bioc", res=2.5)

bio_future <- cmip6_world(
  model = "CNRM-CM6-1", ## modelling method
  ssp   = "245", ## "Middle of the Road" scenario
  time  = "2041-2060", # time period
  var   = "bioc", # just use the bioclim variables
  res   = 2.5
)

# Crop to domain 
bbox <- ext(bb)

bio_current <- crop(bio_current, bbox)
bio_future <- crop(bio_future, bbox)
```

```{r standardize raster layer names, echo = FALSE}
simplify_names <- function(x){
    paste0('bio_', sprintf("%02d", as.numeric(gsub('\\D+','', names(x)))))
}

names(bio_current) <- gsub('^.*5m_', '', names(bio_current))
names(bio_future) <- gsub('^.*2060_', '', names(bio_future))

names(bio_current) <- simplify_names(bio_current)
names(bio_future) <- simplify_names(bio_future)

hemi <- select(hemi) 

rm(simplify_names)
```


## fit the model
Here we fit as SDM using `bayesianSDM`. 
For arguments, it requires occurrence data `x`, a raster stack of `predictors`, a `quantile_v` offset used to create pseudo-absence data, and a `planar_proj` used to calculate distances between occurrence data and possible pseudo-absence points.

```{r Create Species Distribution Model, message=F, warning=F, eval = F}
sdModel <- bayesianSDM( 
  x = hemi,
  pca_predictors = FALSE,
  predictors = bio_current,
  quantile_v = 0.025,
  planar_proj = 5070, 
  resample = TRUE,
  feature_selection = 'ffs',
  min_ffs_var = 5, 
  fact = 3, 
  silent = 2 # print fewer messages during model fit.
  )
```

```{r save model, eval = F}
#setwd('~/Documents/assoRted/safeHavens/vignettes')
#sdModel$RasterPredictions <- terra::wrap(sdModel$RasterPredictions)
#sdModel$Predictors <- terra::wrap(sdModel$Predictors)
#sdModel$RasterPredictions_sd <- terra::wrap(sdModel$RasterPredictions_sd)
#sdModel$AOA <- terra::wrap(sdModel$AOA)
#saveRDS(sdModel, file.path('..', 'inst', 'extdata', 'BayesSDM.rds'))
```

Load a run of the model described above for the sake of this vignettes processing speed.

```{r load pre-computed sdmodel}
sdModel <- readRDS(
  file.path(system.file(package="safeHavens"), 'extdata',  'BayesSDM.rds')
  )

sdModel$RasterPredictions <- terra::unwrap(sdModel$RasterPredictions)
sdModel$Predictors <- terra::unwrap(sdModel$Predictors)
sdModel$RasterPredictions_sd <- terra::unwrap(sdModel$RasterPredictions_sd)
sdModel$AOA <- terra::unwrap(sdModel$AOA)
```

Under the hood, this function uses `brms`, which dispatches to Stan, to fit a Bayesian hierarchical model (GLMM), where the distances between occurrence records are treated as fixed effects to reduce the effects of spatial autocorrelation and increase the interpretability of the model's parameters.

### explore the output

Plot the predictions 
```{r Explore SDM output - Different alpha}
sdm_td <- sdModel$TrainData

par(mfrow = c(1, 2))
plot(sdModel$RasterPredictions, main = 'mean prediction')
points(vect(sdm_td[sdm_td$occurrence==1,]),col = 'black', cex = 0.5)
points(vect(sdm_td[sdm_td$occurrence==0,]),col = 'red', cex = 0.5)

plot(sdModel$RasterPredictions_sd, main = 'sd prediction')
points(vect(sdm_td[sdm_td$occurrence==1,]),col = 'black', cex = 0.5)
points(vect(sdm_td[sdm_td$occurrence==0,]),col = 'red', cex = 0.5)
```

The brms:: model summary can be accessed from the sdModel list.
```{r general moel overview}
sdModel$Model
```

rHat < 1.01 is ideal; < 1.05 is acceptable. Both ESS should be greater than 400; increase iterations to achieve this.
Divergence of 0 is ideal; any divergence is a flag for a possibly misspecified model.

Diagnostics can be accessed from a list, and the stan website has some information on them here, as well as warnings (as may be depicted above) https://mc-stan.org/misc/warnings/

Similar to `elasticSDM`, we also have an area of applicability and other surfaces from CAST.
```{r plot aoa}
plot(sdModel$AOA)
```

```{r check diagnostics}
sdModel$Diagnostics
```

To assess how well the model predicts the suitable habitat for the hold-out sample, the confusion matrix can be obtained from the test and train splits.
While balanced accuracy for these is generally low, as these models are not intended for prediction, so much as understanding the relationships between environmental variables driving the species distribution, the results show the model was able to learn some important characteristics of the species.

```{r Explore SDM output - Confusion Matrix}
knitr::kable(sdModel$ConfusionMatrix$byClass)
```

Observe how the selected model predicts the state of the test data.
Note that these accuracy results are slightly higher than those from the CV folds.
Understand that this is not a bug; CV folds test on their own holdouts, while this is a new holdout set.
The main reason the confusion matrix results are likely to be higher is due to spatial autocorrelation, which is not addressed in a typical ‘random split’ of test and train data.
Minimise the effects of spatial autocorrelation on the model by using CAST for spatially informed cross-validation.

Treat the CVStructure output as a more realistic assessment.

### binarize the output

The Bayesian SDM is post-processed using the same `PostProcessSDM` function as the elasticSDM workflow.

```{r Threshold the SDM output}
threshold_rasts <- PostProcessSDM(
  rast_cont = sdModel$RasterPredictions, 
  test = sdModel$TestData,
  train = sdModel$TrainData,
  planar_proj = 5070,
  thresh_metric = 'equal_sens_spec', 
  quant_amt = 0.2
  )
knitr::kable(threshold_rasts$Threshold)
```

We can compare the results of applying this function side by side using the function's output.
```{r Compare Threshold Results}
terra::plot(threshold_rasts$FinalRasters)
```

### rescale predictor variables
 
The `bayesianSDM` requires a different function from elasticSDM for creating weighted surfaces, with the suffix `RescaleRasters_bayes`.
```{r Rescale Predictor Variables}
rr <- RescaleRasters_bayes(
  model = sdModel$Model,
  predictors = sdModel$Predictors, 
  training_data = sdModel$TrainData, 
  pred_mat = sdModel$PredictMatrix,
  include_uncertainty = TRUE
  )

terra::plot(rr$RescaledPredictors)
```

An optionally returned layer (via `include_uncertainty` = TRUE) depicts areas with greater uncertainty towards the prediction.
```{r plot coef uncertainty layer}
plot(subset(rr$RescaledPredictors, 'coef_uncertainty'))
```

We can also show the variables that contributed the most to the sdm models' predictions, along with their signs.
```{r Show beta Coefficients from model}
knitr::kable(
  rr$BetaCoefficients[ 
    order(abs(rr$BetaCoefficients$Estimate),decreasing = T), 
    ])
```

To delineate putative seed collection zones, we use the `PosteriorCluster` function.
It will use the raster surfaces of the mean beta coefficient importance.

```{r posterior clustering}
obby <- PosteriorCluster(
  model = sdModel$Model,
  predictors = subset(rr$RescaledPredictors, 'coef_uncertainty', negate = TRUE),
  f_rasts = threshold_rasts$FinalRasters,
  lyr = 'Supplemented',
  pred_mat = sdModel$PredictMatrix,
  training_data = sdModel$Train,
  planar_proj = 5070,
  coord_wt = 0.5,
  n_pts = 5000
)
obby$KNNModels
plot(obby$Geometry)
plot(obby$SummaryRaster)
```

Using Bayesian posteriors and clustering allows us to identify areas with higher and lower uncertainty in clustering assignments.
Areas with higher proportions had higher colocation scores than those with lower proportions.

```{r plot stability raster with sample points}
par(mfrow=c(1,1))
plot(obby$SummaryRaster$stability)
points(obby$SamplePoints, cex = 0.5)
```

As with the other sampling functions, spatial polygons are returned for use with other applications.

```{r show posterior clusters, warning=F}
bmap +
  geom_sf(data = obby$Geometry, aes(fill = factor(ID))) + 
  labs(fill = 'Cluster')
```

Interested users can see what the second-most and third-most stable clustering scenarios would look like via the following ranks.
Do note that the renumbering of clusters from NW to SE (left to right, as many language groups read books) has not yet occurred on these rasters, so the numbers differ from the consensus surface.

```{r view top 3 ranks side by side}
par(mfrow = c(1, 3))
plot(obby$SummaryRaster$consensus)
plot(obby$RankRaster$rank2_final)
plot(obby$RankRaster$rank3_final)
par(mfrow = c(1,1))
```

### save results 

writeSDMresults can also be used to save most of the core SDM output from the Bayesian workflow.

```{r Save SDM results, eval = F}
bp <- '~/Documents/assoRted/StrategizingGermplasmCollections'

writeSDMresults(
  cv_model = sdModel$CVStructure, 
  model = sdModel$Model, 
  cm = sdModel$ConfusionMatrix, 
  coef_tab = rr$BetaCoefficients, 
  f_rasts = threshold_rasts$FinalRasters,
  thresh = threshold_rasts$Threshold,
  file.path(bp, 'results', 'SDM'), 'hemi_test')

# we can see that the files were placed here using this. 
list.files( file.path(bp, 'results', 'SDM'), recursive = TRUE )
```


## predictive provenance



```{r}

future_rescaled <- projectClustersBayes(
  bSDM_object = sdModel,
  posterior_clusters = obby,
  future_predictors = bio_future,
  current_predictors = bio_current,
  threshold_rasts = threshold_rasts,
  planar_proj = "epsg:5070",
)

bmap +
  geom_sf(data = future_rescaled$clusters_sf, aes(fill = factor(ID))) + 
  labs(fill = 'Cluster')

```




```{r}
train_df <- obby$KNNModels$KNN_Cluster$trainingData
env_vars <- setdiff(names(train_df), c(".outcome", 'coord_x_w', "coord_y_w"))
env_vars <- intersect(env_vars, names(fut_df))

# Build combined long-format data frame
train_long <- train_df[, env_vars] |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  dplyr::mutate(source = "train (current)")

future_long <- fut_df[stats::complete.cases(fut_df), env_vars] |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  dplyr::mutate(source = "future")

combined <- dplyr::bind_rows(train_long, future_long)

ggplot(combined, aes(x = value, fill = source, colour = source)) +
  geom_density(alpha = 0.35, linewidth = 0.6) +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  scale_fill_manual(values  = c("train (current)" = "#2196F3", "future" = "#FF5722")) +
  scale_colour_manual(values = c("train (current)" = "#1565C0", "future" = "#BF360C")) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "top",
    strip.text       = element_text(size = 8, face = "bold"),
    axis.text        = element_text(size = 7),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title  = "Feature distribution: KNN training (current) vs future rescaled",
    x      = NULL, y      = "density", fill   = NULL, colour = NULL
  )

```

### wrapping up 

While the Bayesian hierarchical GLM does not produce excellent predictions of suitable habitat, it has several advantages.
  1) It can incorporate spatial smoothers (splines) from mcgv to reduce the effects of spatial autocorrelation on model parameter estimates.
  2) It provides uncertainty around the contribution of each selected predictor (i.e. many predictors can be weighed differently to get similar results)

```{r Clean up SDM variables, warning=FALSE, echo = FALSE}
rm(rr, predictors, files, sdModel, bp)
```
